\documentclass[12pt, a4paper, twoside, openright]{book}

\usepackage{vuwthesis} % sets up some local things, mostly the front page


\begin{document}
\chapter*{Acknowledgements}\label{C:ack}

Thanks are due first and foremost to my supervisor, Professor Shaun Hendy, for the opportunity to study fluid mechanics, and the freedom to do it my way.
I would like to acknowledge Professor John Lekner (co-supervisor) for early inspiration in the art of theoretical physics.

The mathematician Dr Xing-You ``Philip" Zhang suggested the use of the homogenization technique; this work has benefited immeasurably as a result.  Philip patiently helped me learn homogenization and tease out the connection between the maths and the physics.  Thank you, Philip.

The entirety of the work was carried out in the (now defunct) Applied Maths team at (the now defunct) Industrial Research Limited.  I wish to thank the Applied Maths team for the resources and hospitality provided to me. I would like to acknowledge Peter McGavin for computer support, the most-recent leader John Burnell, and in particular former leader Graham Weir as an outstanding exemplar of skill in applied mathematics. 
I particularly wish to thank Robin Willink (formerly of Applied Maths), for stimulating discussions on the meaning and interpretation of models.

I have appreciated the company of many students, interns and workers along the way, including Brent, Peter Z, Srikanth, Amanda, Rebecca, Elf, Ben, Geoff the Ref, Keoni, Dion, Nicola, Catriona, Bryne, Udbhav, Kirsten, Donal, Julie, Simon, James, Gregor, Wayne, and various others.

\clearpage
Particular mention must be made of a few fellow students who have now graduated:  my office-mate Dr Sione Paea, for the huge hugs, Dr Dmitri Schebarchov, who kindly read a partial draft of this thesis (thanks D, that helped a lot), Dr Shrividya Ravi, for the conversation and creamy balls, and Dr Krista Steenbergen, for her intensity and huge heart (bigger even than her mouth).

Extra-special feathery felicity to Christine Roseveare, for everything.
\end{document}
